\textbf{Evaluating RISC-V Enclaves:}\\ Security Without the Slowdown?

### 2025-09-02 draft_presentation.tex
improved slides

@@

% Keystone 
\begin{frame}{Why Keystone?}
\begin{itemize}
    \item Modular, open-source TEE for RISC-V.
    \item Uses Physical Memory Protection (PMP) and minimal Security Monitor for isolation.
    \item Low Trusted Computing Base (TCB) for security and flexibility.
    \item Optimized for secure computation in untrusted environments.
\end{itemize}
\begin{block}{Key Concept}
\textit{Low TCB reduces attack surfaces while enabling high flexibility and performance in enclave-based systems.}
\end{block}
\end{frame}

@@

% Keystone vs Other TEEs
\begin{frame}{Keystone vs. Other TEEs}
\begin{figure}[htbp]
\centering
\caption{Feature comparison of major TEEs.}
\label{fig:tee-features}
\renewcommand{\arraystretch}{1.3} % more row height
\setlength{\tabcolsep}{5pt}       % more column padding
\rowcolors{2}{gray!5}{white}      % alternating row colors

\begin{tabular}{@{}lcccc@{}}
\toprule
\rowcolor{gray!20}
\textbf{Feature} & \textbf{Keystone} & \textbf{SGX} & \textbf{TrustZone} & \textbf{SEV} \\
\midrule
Open Source            & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{red!15}No & \cellcolor{red!15}No \\
Dynamic Enclaves       & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{yellow!15}Limited & \cellcolor{red!15}No \\
Hardware Dependency    & \cellcolor{green!15}RISC-V & \cellcolor{red!15}Intel & \cellcolor{red!15}ARM & \cellcolor{red!15}AMD \\
TCB Size               & \cellcolor{green!15}Low & \cellcolor{yellow!15}Medium & \cellcolor{red!15}High & \cellcolor{yellow!15}Medium \\
Flexible Configurations& \cellcolor{green!15}High & \cellcolor{red!15}Low & \cellcolor{red!15}Low & \cellcolor{yellow!15}Medium \\
\bottomrule
\end{tabular}
\end{figure}

@@

\begin{frame}{Benchmarking Tools}
    \small
    \begin{itemize}
        \item \textbf{Dhrystone:} Measures CPU performance (DMIPS).
        \item \textbf{CoreMark:} Focuses on integer performance, used in embedded systems benchmarking.
        \begin{block}{Execution Modes}
        \begin{itemize}
            \item \textbf{Native Execution:} Benchmark runs as a normal process.
            \item \textbf{Enclave Execution:} Benchmark runs within a Keystone enclave.
        \end{itemize}
        \end{block}
    \end{itemize}
    %\vspace{0.5em}
    %\uncover<1->{\textbf{Key Focus:} These tools provide insights into both CPU-bound and system-level performance.}
\end{frame}

%Results
\section{Synthesis of Key Performance Trends and Hardware Bottlenecks}
\label{sec:synthesis}

@@

\begin{frame}{Key Takeaways from Performance Trends}
\small
\begin{block}{Main Findings}
\begin{itemize}
    \item Enclave performance suffers primarily in \textbf{memory-bound workloads} due to security checks and context switching.
    \item \textbf{PMP entry limitations} restrict scalability in multi-enclave setups.
    \item Increasing hardware resources does not linearly translate to performance gains due to system-level overheads.
\end{itemize}
\end{block}

\centering
\textbf{Key Insight:} Optimizing both hardware resources and system-level configurations is essential for scalable and efficient enclave execution.
\end{frame}

\begin{frame}{Why Misalignment Bloats PMP Usage}
\small
\vfill

% Problem
\only<1>{
\begin{block}{Problem}
Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
\end{block}
}

% Cause
\only<2>{
\begin{block}{Problem}
Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
\end{block}
\vspace{0.5em}
\begin{block}{Cause}
When a memory region is \textbf{misaligned}:
\begin{itemize}
  \item NAPOT encoding forces padding to the next power‑of‑two boundary.
  \item Padding areas consume PMP coverage even if they are not actual data.
\end{itemize}
\end{block}
% \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Problem → Cause
}

% Effect
\only<3>{
\begin{block}{Cause}
When a memory region is \textbf{misaligned}:
\begin{itemize}
  \item NAPOT encoding forces padding to the next power‑of‑two boundary.
  \item Padding areas consume PMP coverage even if they are not actual data.
\end{itemize}
\end{block}
\vspace{0.5em}
\begin{block}{Effect}
\begin{itemize}
  \item A single logical region may require \textbf{multiple PMP entries}.
  \item Fragmentation increases as more regions are misaligned.
\end{itemize}
\end{block}
% \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Cause → Effect
}

% Impact
\only<4>{
\begin{block}{Effect}
\begin{itemize}
  \item A single logical region may require \textbf{multiple PMP entries}.
  \item Fragmentation increases as more regions are misaligned.
\end{itemize}
\end{block}
\vspace{0.5em}
\begin{block}{Impact}
\begin{itemize}
  \item Higher PMP entry count $\Rightarrow$ fewer entries left for other regions.
  \item Limits scalability and flexibility of memory protection.
  \item Wastes address space and complicates security policy.
\end{itemize}
\end{block}
}

% Key takeaway
\only<5>{
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1, transform shape,font=\small]

@@

% === Region 1: contiguous, 1 PMP entry ===
\node[anchor=south] at (1.375,2.1) {\scriptsize 128\,KB aligned};
%\draw[pattern=north east lines,pattern color=black] (0.25,0) rectangle (0.5,1);
\draw[fill=keystoneblue!60] (0.5,0) rectangle (2.2,1);
%\draw[pattern=north east lines,pattern color=black] (2.2,0) rectangle (2.5,1);
\draw[fill=keystoneblue!60,opacity=0.4] (0.5,1) rectangle (2.2,2) node[pos=.5] {\scriptsize 1 entry};

@@

% Arrows showing cause-effect
%\draw[->,thick] (2.2,0.5) .. controls (2.8,1.5) .. (3.4,2) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
\draw[->,thick] (5.1,0.5) .. controls (5.4,2) .. (6.6,2.5) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};

@@

\vspace{0.4em}
\scriptsize
\textbf{Takeaway:} Align regions to NAPOT boundaries to minimize padding and preserve PMP entries.
\end{figure}
}

\vfill
\end{frame}

@@

% Cloud
\node[box, fill=blue!20] (cloud) {\includegraphics[scale=0.05]{example-image} \\ \textbf{Cloud Server} \\ (Untrusted Environment)};

% Lock / Enclave
\node[box, fill=green!20, right=3cm of cloud] (enclave) {\faLock \\ \textbf{Keystone Enclave} \\ Secure Execution};

@@

% Summary of Findings
\begin{frame}{Key Takeaways}
\begin{itemize}
    \item Keystone offers strong security with minimal overhead for CPU-bound workloads.
    \item Significant overhead for memory bound tasks can be reduced with optimizations.
    \item Customizable settings allow for optimization based on workload type.
\end{itemize}
\end{frame}

>>>>

\begin{frame}{Contributions}
\begin{itemize}
    \item \textbf{Systematic performance evaluation} of Keystone across varied workloads.
    \item \textbf{Identification of hardware bottlenecks} like PMP entry exhaustion and misaligned memory.
    \item \textbf{Actionable optimization guidelines} for developers using Keystone TEEs.
    %\item \textbf{Visual benchmarking workflow} and comparison to alternative TEEs.
\end{itemize}
\end{frame}

@@

% Keystone 
\begin{frame}{From Problem to Solution: Enter Keystone}
\begin{itemize}
    \item Modular, open-source TEE for RISC-V.
    \item Uses Physical Memory Protection (PMP) and minimal Security Monitor for isolation.
    \item Low Trusted Computing Base (TCB) for security and flexibility.
    \item Optimized for secure computation in untrusted environments.
\end{itemize}
\begin{block}{Key Concept}
\textit{Low TCB reduces attack surfaces while enabling high flexibility and performance in enclave-based systems.}
\end{block}
\end{frame}

@@

% Keystone vs Other TEEs
\begin{frame}{Keystone vs. Other TEEs}
\begin{figure}[htbp]
\centering
\renewcommand{\arraystretch}{1.3} % more row height
\setlength{\tabcolsep}{5pt}       % more column padding
\rowcolors{2}{gray!5}{white}      % alternating row colors

\begin{tabular}{@{}lcccc@{}}
\toprule
\rowcolor{gray!20}
\textbf{Feature} & \textbf{Keystone} & \textbf{SGX} & \textbf{TrustZone} & \textbf{SEV} \\
\midrule
Open Source            & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{red!15}No & \cellcolor{red!15}No \\
Dynamic Enclaves       & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{yellow!15}Limited & \cellcolor{red!15}No \\
Hardware Dependency    & \cellcolor{green!15}RISC-V & \cellcolor{red!15}Intel & \cellcolor{red!15}ARM & \cellcolor{red!15}AMD \\
TCB Size               & \cellcolor{green!15}Low & \cellcolor{yellow!15}Medium & \cellcolor{red!15}High & \cellcolor{yellow!15}Medium \\
Flexible Configurations& \cellcolor{green!15}High & \cellcolor{red!15}Low & \cellcolor{red!15}Low & \cellcolor{yellow!15}Medium \\
\bottomrule
\end{tabular}
\caption{Feature comparison of major TEEs.}
\label{fig:tee-features}
\end{figure}

@@

% \begin{frame}{Benchmarking Tools}
%     \small
%     \begin{itemize}
%         \item \textbf{Dhrystone:} Measures CPU performance (DMIPS).
%         \item \textbf{CoreMark:} Focuses on integer performance, used in embedded systems benchmarking.
%         \begin{block}{Execution Modes}
%         \begin{itemize}
%             \item \textbf{Native Execution:} Benchmark runs as a normal process.
%             \item \textbf{Enclave Execution:} Benchmark runs within a Keystone enclave.
%         \end{itemize}
%         \end{block}
%     \end{itemize}
%     %\vspace{0.5em}
%     %\uncover<1->{\textbf{Key Focus:} These tools provide insights into both CPU-bound and system-level performance.}
% \end{frame}

%Results
\section{Key Performance Trends and Hardware Bottlenecks}
\label{sec:synthesis}

@@

% \begin{frame}{Key Takeaways from Performance Trends}
% \small
% \begin{block}{Main Findings}
% \begin{itemize}
%     \item Enclave performance suffers primarily in \textbf{memory-bound workloads} due to security checks and context switching.
%     \item \textbf{PMP entry limitations} restrict scalability in multi-enclave setups.
%     \item Increasing hardware resources does not linearly translate to performance gains due to system-level overheads.
% \end{itemize}
% \end{block}

% \centering
% \textbf{Key Insight:} Optimizing both hardware resources and system-level configurations is essential for scalable and efficient enclave execution.
% \end{frame}

% \begin{frame}{Why Misalignment Bloats PMP Usage}
% \small
% \vfill

% % Problem
% \only<1>{
% \begin{block}{Problem}
% Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
% \end{block}
% }

% % Cause
% \only<2>{
% \begin{block}{Problem}
% Physical Memory Protection (PMP) entries must cover power‑of‑two sized, naturally aligned regions (NAPOT).
% \end{block}
% \vspace{0.5em}
% \begin{block}{Cause}
% When a memory region is \textbf{misaligned}:
% \begin{itemize}
%   \item NAPOT encoding forces padding to the next power‑of‑two boundary.
%   \item Padding areas consume PMP coverage even if they are not actual data.
% \end{itemize}
% \end{block}
% % \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Problem → Cause
% }

% % Effect
% \only<3>{
% \begin{block}{Cause}
% When a memory region is \textbf{misaligned}:
% \begin{itemize}
%   \item NAPOT encoding forces padding to the next power‑of‑two boundary.
%   \item Padding areas consume PMP coverage even if they are not actual data.
% \end{itemize}
% \end{block}
% \vspace{0.5em}
% \begin{block}{Effect}
% \begin{itemize}
%   \item A single logical region may require \textbf{multiple PMP entries}.
%   \item Fragmentation increases as more regions are misaligned.
% \end{itemize}
% \end{block}
% % \draw[->, thick] (0,-0.2) -- (0,-0.8); % optional arrow from Cause → Effect
% }

% % Impact
% \only<4>{
% \begin{block}{Effect}
% \begin{itemize}
%   \item A single logical region may require \textbf{multiple PMP entries}.
%   \item Fragmentation increases as more regions are misaligned.
% \end{itemize}
% \end{block}
% \vspace{0.5em}
% \begin{block}{Impact}
% \begin{itemize}
%   \item Higher PMP entry count $\Rightarrow$ fewer entries left for other regions.
%   \item Limits scalability and flexibility of memory protection.
%   \item Wastes address space and complicates security policy.
% \end{itemize}
% \end{block}
% }

% % Key takeaway
% \only<5>{
% \centering
% \begin{figure}[htbp]
% \centering
% \begin{tikzpicture}[scale=1, transform shape,font=\small]

% % Memory axis
% \draw[thick,->] (0,0) -- (10,0) node[anchor=west]{Memory address};
% \foreach \x/\label in {0/0\,KB,1.25/128\,KB,4.25/192\,KB,8/300\,KB}
%     \draw (\x,0.1) -- (\x,-0.1) node[anchor=north]{\label};

% % === Region 1: contiguous, 1 PMP entry ===
% \node[anchor=south] at (1.375,2.1) {\scriptsize 128\,KB aligned};
% %\draw[pattern=north east lines,pattern color=black] (0.25,0) rectangle (0.5,1);
% \draw[fill=keystoneblue!60] (0.5,0) rectangle (2.2,1);
% %\draw[pattern=north east lines,pattern color=black] (2.2,0) rectangle (2.5,1);
% \draw[fill=keystoneblue!60,opacity=0.4] (0.5,1) rectangle (2.2,2) node[pos=.5] {\scriptsize 1 entry};

% % === Region 2: fragmented, 3 PMP entries ===
% \node[anchor=south] at (4.3,4.1) {\scriptsize 192\,KB misaligned};
% \draw[pattern=north east lines,pattern color=black] (3.5,0) rectangle (3.8,1);
% \draw[fill=warmaccent!60] (3.8,0) rectangle (4.6,1);
% \draw[pattern=north east lines,pattern color=black] (4.6,0) rectangle (5.1,1);
% \draw[fill=warmaccent!60,opacity=0.4] (3.5,1) rectangle (5.1,4) node[pos=.5] {\scriptsize 3 entries};

% % === Region 3: fragmented, 3 PMP entries ===
% \node[anchor=south] at (8.0,4.1) {\scriptsize 300\,KB misaligned};
% \draw[pattern=north east lines,pattern color=black] (6.7,0) rectangle (7.0,1);
% \draw[fill=warmaccent!60] (7.0,0) rectangle (9.1,1);
% \draw[pattern=north east lines,pattern color=black] (9.1,0) rectangle (9.3,1);
% \draw[fill=warmaccent!60,opacity=0.4] (6.7,1) rectangle (9.3,4) node[pos=.5] {\scriptsize 3 entries};

% % Arrows showing cause-effect
% %\draw[->,thick] (2.2,0.5) .. controls (2.8,1.5) .. (3.4,2) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};
% \draw[->,thick] (5.1,0.5) .. controls (5.4,2) .. (6.6,2.5) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};

% % Legend
% \begin{scope}[xshift=0cm,yshift=-1.5cm]
% \draw[fill=keystoneblue!60] (0,0) rectangle (0.4,0.4) node[right=0.5cm] {Contiguous region};
% \draw[fill=warmaccent!60] (5,0) rectangle (5.4,0.4) node[right=0.5cm] {Fragmented region};
% \draw[pattern=north east lines,pattern color=black] (0,-0.6) rectangle (0.4,-0.2) node[right=0.5cm] {Wasted padding};
% \draw[fill=gray!50,opacity=0.4] (5,-0.6) rectangle (5.4,-0.2) node[right=0.5cm] {PMP entry count bar};
% \end{scope}

% \end{tikzpicture}

% \vspace{0.4em}
% \scriptsize
% \textbf{Takeaway:} Align regions to NAPOT boundaries to minimize padding and preserve PMP entries.
% \end{figure}
% }

% \vfill
% \end{frame}

\begin{frame}{PMP Entries vs Max Concurrent Enclaves}
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=10cm,
    height=6cm,
    xlabel={PMP Entries},
    ylabel={Max Concurrent Enclaves},
    xmin=0, xmax=16,
    ymin=0, ymax=16,
    xtick={0,4,8,12,16},
    ymajorgrids,
    grid style=dashed,
    mark options={solid},
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    clip=false,
    cycle list={{teal,mark=*,thick}},
    tick label style={font=\small},
    label style={font=\small\bfseries},
]

% Data plot
\addplot[
    color=teal,
    mark=*,
    thick
] coordinates {
    (0,0) (2,2) (4,4) (6,6) (8,8) (10,8) (12,8) (14,8) (16,8)
};

% Highlight PMP exhaustion area
\fill[red!30,opacity=0.4] (axis cs:8,0) rectangle (axis cs:16,16);

% Annotation inside shaded region
\node[red, align=center, font=\footnotesize\bfseries] at (axis cs:12,12) {PMP Exhaustion \\ Caps at 8 Enclaves};

% Stability boundary line & label
\draw[dashed, thick, color=blue] (axis cs:4,0) -- (axis cs:4,16);
\node[anchor=south, font=\footnotesize\bfseries, color=blue] at (axis cs:4,16) {Stability Boundary};

\end{axis}
\end{tikzpicture}
\caption{Max enclaves increase with PMP entries until exhaustion caps growth.}
\end{figure}
\end{frame}

\begin{frame}{PMP Misalignment: Why It Matters}
\begin{block}{The Problem}
PMP entries must describe power-of-two, naturally aligned regions (NAPOT). Misaligned memory allocations cause unnecessary padding.
\end{block}

\begin{block}{The Effect}
\begin{itemize}
    \item Misalignment causes a single memory region to consume multiple PMP entries.
    \item Fragmentation leads to rapid exhaustion of PMP resources (max = 16 entries).
\end{itemize}
\end{block}

\vspace{1em}
\centering
\textbf{Next: Visualizing the impact →}
\end{frame}

\begin{frame}{PMP Misalignment: Visualized}
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1, transform shape,font=\small]

@@

% === Region 1: contiguous, 1 PMP entry ===
\node[anchor=south] at (1.375,2.1) {\scriptsize 128\,KB aligned};
\draw[fill=keystoneblue!60] (0.5,0) rectangle (2.2,1);
\draw[fill=keystoneblue!60,opacity=0.4] (0.5,1) rectangle (2.2,2) node[pos=.5] {\scriptsize 1 entry};

@@

% Arrows showing cause-effect
\draw[->,thick] (5.1,0.5) .. controls (5.4,2) .. (6.6,2.5) node[midway,above,sloped]{\scriptsize Padding $\Rightarrow$ More entries};

@@

\vspace{0.4em}
\scriptsize
\textbf{Takeaway:} Align regions to NAPOT boundaries to minimize padding and preserve PMP entries.
\end{figure}
\end{frame}

@@

% Cloud
\node[box, fill=blue!20] (cloud) {\faCloud \\ \textbf{Cloud Server} \\ (Untrusted Environment)};

% Lock / Enclave
\node[box, fill=green!20, right=3cm of cloud] (enclave) {\faLock \\ \textbf{Keystone Enclave} \\ Trusted Execution};

@@

% % Summary of Findings
% \begin{frame}{Key Takeaways}
% \begin{itemize}
%     \item Keystone offers strong security with minimal overhead for CPU-bound workloads.
%     \item Significant overhead for memory bound tasks can be reduced with optimizations.
%     \item Customizable settings allow for optimization based on workload type.
% \end{itemize}
% \end{frame}

\begin{frame}{Summary of Findings}
\begin{itemize}
    \item \textbf{Keystone offers low-overhead enclaves} for CPU-bound workloads.
    \item \textbf{Memory-bound performance degrades} due to PMP overhead and context switching.
    \item \textbf{Scalability limited} by fixed PMP entries and misaligned region padding.
    \item \textbf{Performance can be improved} via alignment and workload partitioning strategies.
\end{itemize}
\end{frame}
